Data Specialization for a Staged Lambda Calculus

### 2015-06-27 paper/semantics.tex
changes to semantics

\input{figures/values}

@@

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms [need ``series expression'' citation!], 
as this will be useful in defining our splitting algorithm.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

We provide a small-step semantics which evaluates the first stage subterms
of any \lang\ term.  
This requires notions of values and steps that are specialized to each world.

@@

The statics and dynamics of \lang\ are related by a type safety theorem
comprising the usual progress and preservation lemmas, again annotated by world:
\\{\bf Progress.~} If $\typeswor e A$, then either 
$\stepwor e {e'}$ or $e \done$ or $\lift e {e'}$ (hoisting available only at $w = \bbonem$).
\\{\bf Preservation.~} For normal steps, if $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$. 
For hoisting steps, if $\typesone e A$ and $\lift e {e'}$, then $\typesone {e'} A$.

In the typing rules above, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage variables under which we are evaluating.

@@

We consider a staged program to be a closed term $e$ at world \bbtwo. The first
stage of evaluation is $e\mathbin{\redtwosym} q$, which reduces all code at
\bbonem\ and \bbonep\ inside $e$. The second stage of evaluation is $q
\mathbin{\redsym} u$, or monostaged evaluation of the residual $q$, which yields
a value $u$.

>>>>

@@

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which includes and subsumes the concept of a stage.
Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms, 
with an underline (``$\exv v$'') noting the explicit inclusion of values into general terms.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
$\exv q$, and the second pass further reduces this residual to a value.
Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
Accordingly, this paper concerns itself only with the first pass of evaluation.

Since $\coltwo e A$ may be constructed out of terms at other world, 
our dynamics requires notions of values and steps that are specialized to each world.

@@

The statics and dynamics of \lang\ are related by a type safety theorem
comprising the usual progress and preservation lemmas, again annotated by world.
In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
\\{\bf Progress.~} 
\begin{itemize}
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
{\bf Preservation.~} 
\begin{itemize}
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}

@@
